---
title: "Trialling the *fluxweb* package"
author: "Ben Moon"
date: 2022-02-03
...

# Environment

I use *renv* to maintain a reproducible environment.

```{r initialize_renv}
renv::init()
```

ACtivating the *renv* project should install the required packages that can then be loaded as usual.

```{r load_packages}
library(dplyr)
library(fluxweb)
library(ggplot2)
library(igraph)
library(patchwork)
library(purrr)
library(tibble)
library(tidyr)
```

# Using *fluxweb* #

*fluxweb* requires several data sets:

* a matrix of trophic interactions
* a vector of body masses
* a vector of total biomass
* a vector of the organism type.

There is also a vector of metabolic types that is not required, but can be modified to increase precision.


## A random simulation ##

### The initial data ###


I'll start with a randomly generated example, as this is more fun.

```{r random_matrix_generation, eval = FALSE}
random_trophic_interactions <-
  matrix(sample(c(0, 1), size = 100, replace = TRUE), nrow = 10)
random_trophic_interactions[upper.tri(random_trophic_interactions, diag = TRUE)] <- 0
# saveRDS(random_trophic_interactions, file = "random_trophic_interactions.rds")
```

Here is the matrix I generated, use this to keep it consistent and easier for generating the relative masses below. Groups 9 and 10 are at the bototm of the food web, and 5, 2, and 1 are at the top.

```{r random_matrix}
random_trophic_interactions <- readRDS("random_trophic_interactions.rds")
random_trophic_interactions
#       [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10]
#  [1,]    0    0    0    0    0    0    0    0    0     0
#  [2,]    0    0    0    0    0    0    0    0    0     0
#  [3,]    1    1    0    0    0    0    0    0    0     0
#  [4,]    0    0    1    0    0    0    0    0    0     0
#  [5,]    0    1    1    0    0    0    0    0    0     0
#  [6,]    0    0    1    0    1    0    0    0    0     0
#  [7,]    0    1    0    0    1    0    0    0    0     0
#  [8,]    0    1    1    1    1    0    0    0    0     0
#  [9,]    0    0    1    0    1    0    1    1    0     0
# [10,]    0    0    0    1    0    1    1    0    0     0
random_web_layout <-
  matrix(c(2, 1, 2, 1, 3, 3, 2, 1, 1, 3,
           4, 3, 3, 2, 2, 1, 2, 1, 0, 0), ncol = 2)
```

For each of the ten trophic groups/species, I create some random body masses but add some patterning so that the prey tends to be smaller than its predators. Similarly for the biomass of each group.

```{r random_masses, echo = FALSE options}
random_body_masses <- c(5000, 1000, 1000, 500, 1000, 100, 500, 100, 1, 10)
random_biomasses <- c(10, 10, 5, 50, 10, 5, 10, 50, 100, 100) * 100
```

Finally, I include the organism type in a vector.

```{r random_organism_type}
random_organism_type <- c(rep("animal", 8), rep("plant", 2))
random_metabolic_type <- c(rep("endo.vert", 2), rep("ecto.vert", 3), rep("inv", 5))
```

I show the layout of this using *igraph* with node sizes scaled to biomass.

```{r igraph_simulated_example}
plot(graph_from_adjacency_matrix(random_trophic_interactions, mode="directed"), layout = random_web_layout, vertex.size = random_biomasses * 0.01)
```

### Calculating efficiencies and losses ###

The losses and the efficiencies are the values that fluxing requires to stabilize the network. *Losses* are the energy spent on metabolisms (= metabolic rate) that is thus lost to organisms further up the food chain. *Efficiencies* incorporates the efficiency of eating and how much energy is lost from that.

Metabolic rate \(X\) is estimated with the exponential relationship

\[ X = a M^{b} \]

where \(M\) is the body mass. The following values are used for different metabolic groups.

```{r metabolic_types_losses}
metabolic_losses <-
  data.frame(
    ecto.vert = c(18.18, -0.29),
    endo.vert = c(19.5,  -0.29),
    inv       = c(18.18, -0.29)
  )

random_losses <-
  list(mass = random_body_masses, metabolism = random_metabolic_type) |>
  pmap_dbl(
    function (mass, metabolism, rates) {
      rates[1, metabolism] * mass ^ rates[2, metabolism]
    },
    rates = metabolic_losses
  )
```

Feeding efficiencies are similarly included using the following values.

```{r random_efficiencies}
organisms_efficiency <-
  data.frame(animal = 0.906, plant  = 0.545, detritus = 0.158)
random_efficiencies <- random_organism_type |>
  map_dbl(function (type, efficiency) { efficiency[1, type] }, efficiency = organisms_efficiency )
```

### Fluxing ###

I calculate the fluxes through the web with the `fluxing` function.

```{r fluxing}
random_fluxes <- fluxing(random_trophic_interactions, random_biomasses, random_losses, random_efficiencies)
```

It's as simple as that.

### Fluxes to function ###

The vignette in *fluxweb* adds the next step of converting the fluxes into ecosystem function. The fluxes calculated are the losses of energy from the resource nodes, i.e. the energy that doesn't make it to the next step in the chain.

```{r identify_functions}
basals <- colSums(random_fluxes) == 0
plants <- basals

random_functions <- data.frame(
  herbivory = sum(rowSums(random_fluxes[plants, ])),
  carnivory = sum(rowSums(random_fluxes[!basals, ])),
  total     = sum(random_fluxes)
)
```

The functions are identified as the sums of the outgoing fluxes for the consumed items. For example, herbivory is the flux value outgoing from plants.

```{r random_plot_fluxes}
random_functions |>
  pivot_longer(everything(), names_to = "Function", values_to = "Fluxes") |>
  ggplot(aes(x = Function, y = Fluxes)) +
    geom_col()
```

### Food web stability ###

Calculating the stability value is not in the vignette, but I've gathered this from Sander et al. (2021, https://doi.org/10.1126/science.abf5787). I estimate the stability based on eigenvalues for the food web, the more negative this value the more stable the food web. This also requires growth rates, which apparently applies onto to invertebrate groups; taxa 9, 10.

```{r random_stability}
random_growth_rates <- c(rep(NA, 8), rep(0.5, 2))
random_stability <- stability.value(random_fluxes, random_biomasses, random_losses, random_efficiencies, random_growth_rates, ef.level = "pred")
```

This value is only really useful when you have things to compare it to.


## A Pokémon-based example ##

Now for a 'real world' example. Kanto Route 8 passes between Lavender Town to Saffron City and features only one patch of long grass that contains all the Pokémon, making it somewhat easier to catch 'em all. There are 12 species of Pokémon found on this route as I'm combining occurrences from Red, Blue, and Yellow into one: Pidgey, Pidgeotto, Rattata, Ekans, Sandshrew, Vulpix, Jigglypuff, Meowth, Mankey, Growlithe, Abra, and Kadabra.

First I'm going to build up the trophic interactions.

```{r pokemon_trophic_interactions}
pokemon_names <- c("Pidgey", "Pidgeotto", "Rattata", "Ekans", "Sandshrew", "Vulpix", "Jigglypuff", "Meowth", "Mankey", "Growlithe", "Abra", "Kadabra", "Grass", "Berries")
pokemon_trophic_interactions <-
  data.frame(
    eaten = c("Pidgey", "Pidgeotto", "Rattata", "Meowth", "Mankey", "Pidgey", "Rattata", "Sandshrew", "Pidgey", "Rattata", "Rattata", "Abra", "Rattata", "Ekans", "Sandshrew", "Grass", "Berries", "Berries", "Grass", "Berries", "Grass", "Grass", "Berries", "Jigglypuff", "Berries", "Berries", "Jigglypuff"),
    eats  = c(rep("Growlithe", 5), rep("Ekans", 3), rep("Meowth", 2), rep("Vulpix", 2), rep("Pidgeotto", 3), rep("Pidgey", 2), "Rattata", rep("Sandshrew", 2), "Jigglypuff", rep("Mankey", 3), "Abra", rep("Kadabra", 2))
  )
pokemon_trophic_network <- graph.data.frame(pokemon_trophic_interactions)
pokemon_trophic_matrix <- as.matrix(get.adjacency(pokemon_trophic_network))
```

And now to build up the biomasses. I've decided to add some variation with simulation. To do this I simulate walking through the long grass for one hour (at one-step-per-second) and encountering Pokémon (then immediately fleeing) as I go using the overall encounter rates (20/255 by default) and relative rates between Pokémon species. I also take the respective body masses and use a normal distribution to 'encounter' a variety of Pikeman sizes.

```{r simulate_pokemon_encounters}
encounter_rate <- 20 / 255

route_8_pokemon <- data.frame(
  pokemon = pokemon_names[1:12],
  encounter_rate = c(35, 10, 15, 20, 20, 20, 10, 25, 25, 20, 20, 5) / 100,
  adjusted_encounter_rate = encounter_rate / sum(encounter_rate),
  mass = c(1.8, 30, 3.5, 6.9, 12.0, 9.9, 5.5, 4.2, 28.0, 19.0, 19.5, 56.5) * 1000
)

encounter_simulation <-
  imap_dfr(1:3600, function(.x, .y) {
    r_num <- sample(1:255, size = 1) / 255
    if (r_num < encounter_rate) {
      found_pokemon <- sample(route_8_pokemon$pokemon, size = 1, prob = route_8_pokemon$adjusted_encounter_rate)
      found_mass <- route_8_pokemon[route_8_pokemon$pokemon == found_pokemon, "mass"]
      pokemon_mass <- rnorm(1, found_mass, sd = found_mass / 10)
      data.frame(pokemon = found_pokemon, mass = pokemon_mass)
    }
})
```

I show the numbers and ranges of Pikeman encountered here.

```{r plot_pokemon_encounters}
encountered_masses_plot <-
encounter_simulation |>
  ggplot(aes(x = mass, colour = pokemon, fill = pokemon)) +
    geom_histogram() +
    facet_wrap(vars(pokemon), ncol = 2) +
    theme(legend.position = "none")

pokemon_body_mass <- encounter_simulation |>
  group_by(pokemon) |>
  summarize(mass = mean(mass))

encountered_biomass <-
  encounter_simulation |>
    group_by(pokemon) |>
    summarize(biomass = sum(mass))

encountered_biomass_plot <-
  encountered_biomass |>
    ggplot(aes(y = pokemon, x = biomass, colour = pokemon, fill = pokemon)) +
      geom_col() +
      theme(legend.position = "none")

encountered_masses_plot + encountered_biomass_plot
```

And the relevant metabolism categories for each Pokémon.

```{r pokemon_metabolism}
pokemon_organism_type <- c(rep("animal", 12), rep("plant", 2))
pokemon_metabolic_type <- c(rep("endo.vert", 12), rep("inv", 2))
pokemon_metabolic_type[c(4, 11, 12)] <- "ecto.vert"
pokemon_metabolic_type[7] <- "inv"
```

The last things to add are the amounts of grass and berries found. Grass is in 49 squares, so with an estimate of the amount per-square I can add that in. Berries are another random encounter, but I'll include that like grass and they may be scattered and found more often than just on trees.

```{r grass_amount}
plant_biomass <- data.frame(Grass = rnorm(49, 398, 55), Berries = rnorm(49, 340 * 5, 75)) |>
  pivot_longer(everything(), names_to = "pokemon", values_to = "mass")
plant_body_mass <- c(10, 5)
summ_plant_biomass <- plant_biomass |>
  group_by(pokemon) |>
  summarize(biomass = mean(mass))

pokemon_body_mass <- pokemon_body_mass |>
  bind_rows(tibble(pokemon = c("Grass", "Berries"), mass = plant_body_mass))
pokemon_biomass <- encountered_biomass |>
  bind_rows(summ_plant_biomass)
```

(Grass value from <https://www.researchgate.net/figure/biomass-production-in-grams-per-square-meter-separated-by-sites-dominated-by-grass-or_tbl2_233861647>, berries value from <https://www.researchgate.net/publication/317780169_Berry_production_drives_bottom_-_up_effects_on_body_mass_and_reproductive_success_in_an_omnivore>.)

```{r pokemon_losses}
pokemon_losses <-
  list(mass = pokemon_body_mass$mass, metabolism = pokemon_metabolic_type) |>
  pmap_dbl(
    function (mass, metabolism, rates) {
      rates[1, metabolism] * mass ^ rates[2, metabolism]
    },
    rates = metabolic_losses
  )
```

Feeding efficiencies are similarly included using the following values.

```{r random_efficiencies}
organisms_efficiency <-
  data.frame(animal = 0.906, plant  = 0.545, detritus = 0.158)
pokemon_efficiencies <- pokemon_organism_type |>
  map_dbl(function (type, efficiency) { efficiency[1, type] }, efficiency = organisms_efficiency )
```

And finally resolve the fluxes within the web.

```{r pokemon_fluxes}
pokemon_fluxes <- fluxing(pokemon_trophic_matrix, pokemon_biomass$biomass, pokemon_losses, pokemon_efficiencies)
```

Let's see what this looks like in out network, scaling the edges by the flux size.

```{r pokemon_scaled_network}
pokemon_network_layout <-
  matrix(c(0, 1, 2, 3, 5, 4, 6, 7, 3, 5, 8, 2, 4, 6,
           1, 2, 1, 2, 2, 1, 1, 2, 0, 0, 1, 3, 3, 3), ncol = 2)
pokemon_flux_network <- graph.adjacency(pokemon_fluxes, mode = "directed", weighted = TRUE)
plot(pokemon_flux_network, edge.width = E(pokemon_flux_network)$weight / 50000, layout = pokemon_network_layout)
```

And the relative function of the different groups.

```{r pokemon_functions}
basals <- colSums(pokemon_fluxes) == 0
plants <- basals

pokemon_functions <- data.frame(
  herbivory = sum(rowSums(pokemon_fluxes[plants, ])),
  carnivory = sum(rowSums(pokemon_fluxes[!basals, ])),
  total     = sum(pokemon_fluxes)
)
```

The functions are identified as the sums of the outgoing fluxes for the consumed items. For example, herbivory is the flux value outgoing from plants.

```{r pokemon_plot_fluxes}
pokemon_functions |>
  pivot_longer(everything(), names_to = "Function", values_to = "Fluxes") |>
  ggplot(aes(x = Function, y = Fluxes)) +
    geom_col()
```

```{r random_stability}
pokemon_growth_rates <- c(rep(NA, 12), rep(0.5, 2))
pokemon_growth_rates[7] <- 0.5
pokemon_stability <- stability.value(pokemon_fluxes, pokemon_biomass$biomass, pokemon_losses, pokemon_efficiencies, pokemon_growth_rates, ef.level = "pred")
```
